---
title: "jmseq: joint modelling of longitudinal and time-to-event data"
author: Paul McKeigue
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{jmseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include=FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  cache      = TRUE,
  cache.path = "jmseq_cache/",
  fig.width  = 7,
  fig.height = 4
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

The computation in this vignette is performed by the Python package `jmseq`, called from R via the [`reticulate`](https://rstudio.github.io/reticulate/) package.  Results are returned to R as ordinary data frames and lists, and displayed using `ggplot2` and `kableExtra`.

The `jmseq` Python package fits joint models of longitudinal biomarker trajectories and time-to-event outcomes.  The longitudinal submodel is a continuous-time linear Gaussian state space model (LGSSM) estimated by MAP inference using a JAX Kalman filter and NumPyro.  The survival submodel is a Poisson hazard model that takes the Kalman-filtered states as time-varying covariates.

This vignette demonstrates 4-fold cross-validated prediction of mortality on the Mayo Clinic primary biliary cirrhosis (PBC) dataset, comparing a linear mixed model (random slopes only) to a fuller SDE model with autoregressive drift and diffusion.

# Running Python from R with reticulate

The `reticulate` package embeds a Python session inside the R session.  Three functions are used throughout this vignette:

- **`import("module")`** — imports a Python module and returns an R object whose `$` operator maps to Python attribute access.  For example, `jm <- import("jmseq")` makes every public function in `jmseq` available as `jm$function_name(...)`.
- **`r_to_py(x)`** — converts an R object to its Python equivalent.  R data frames become `pandas.DataFrame` objects; R vectors become numpy arrays.
- **`py_to_r(x)`** — converts a Python object back to R.  `pandas.DataFrame` → R data frame; Python lists and dicts → R lists.

Python objects returned by functions called via `$` are converted to R automatically when they are standard types (data frames, numeric scalars, strings, lists).  No explicit `py_to_r()` call is needed in most cases.

# Setup

```{r setup, cache=FALSE}
library(reticulate)
library(ggplot2)
library(kableExtra)
library(data.table)

## Specify which Python executable to use.
## The jmseq Python package must be installed in this Python environment
## (pip install -e . from the repository root).
use_python("/usr/bin/python3", required = TRUE)

ggplot2::theme_set(ggplot2::theme_light(base_size = 9))
```

```{r imports, cache=FALSE}
## import() is the reticulate equivalent of Python's import statement.
## After these calls, jm$split_surv_long(...) calls jmseq.split_surv_long(...),
## and so on for every public function in the package.
jm <- import("jmseq")
np <- import("numpy")
pd <- import("pandas")
```

# Dataset

The PBC dataset comprises 1945 longitudinal measurements of two biomarkers (log-bilirubin and albumin, both standardised) on 312 individuals followed for mortality up to 15 years from baseline.  Time-invariant covariates are sex, age at baseline, and treatment assignment.

The data are loaded from the R package data object `pbc`, then converted to plain R data frames before being passed to Python.

```{r data, cache=FALSE}
## Path relative to the vignette directory (jmseq/vignettes/ → jmseq/data/)
load("../data/pbc.rda")

## Convert the treatment factor to a binary integer (D-penicillamine = 1).
## The Python pipeline works with numeric covariates only.
dataSurv_r <- as.data.frame(pbc$dataSurv)
dataSurv_r$trt_binary <- as.integer(dataSurv_r$trt == "D-penicil")
dataSurv_r <- dataSurv_r[, c("id", "sex", "agebaseline", "trt_binary",
                              "Time.cens", "event")]

dataLong_r <- as.data.frame(pbc$dataLong[, c("id", "Time",
                                              "logBili", "albumin")])

biomarkers     <- c("logBili", "albumin")
timeinvar_surv <- c("sex", "agebaseline", "trt_binary")

cat(sprintf("dataSurv: %d individuals, %d events\n",
            nrow(dataSurv_r), sum(dataSurv_r$event)))
cat(sprintf("dataLong: %d observations\n", nrow(dataLong_r)))
```

## Preprocessing: interval splitting

Follow-up time is split into intervals of at most 0.25 years so that the Poisson hazard model approximates a continuous-time process.  Intervals containing a biomarker observation are flagged so the Kalman filter performs an update step; empty intervals carry only a prediction step.

The R data frames are passed to `jm$split_surv_long()` using `r_to_py()`.  The function returns a Python `pandas.DataFrame`, which reticulate automatically converts to an R data frame when it is assigned to an R variable.

```{r split}
## r_to_py() converts R data.frames to pandas DataFrames.
surv_py <- r_to_py(dataSurv_r)
long_py <- r_to_py(dataLong_r)

## jm$split_surv_long() is the Python function jmseq.split_surv_long().
## The returned pandas DataFrame is automatically coerced to an R data frame.
split_df <- jm$split_surv_long(
  surv_py,
  long_py,
  max_interval   = 0.25,
  timeinvar_surv = timeinvar_surv,
  biomarkers     = biomarkers
)

dataSurv_split <- split_df[c("id", "Time", "tstop", "event", timeinvar_surv)]
dataLong_split <- split_df[c("id", "Time", "tstop", biomarkers)]

cat(sprintf("After splitting: %d rows, max T per individual = %d\n",
            nrow(split_df),
            max(table(py_to_r(split_df["id"])))))
```

# Model variants

We compare two LGSSM variants for the biomarker trajectories.  The five available variants are pre-defined in `jmseq.VARIANTS`; `make_config()` selects one by name and sets the state dimension `p`.  In R, integer arguments to Python functions must be written as `2L` rather than `2` to avoid being passed as floating-point.

```{r configs}
## make_config() returns a Python LGSSMConfig dataclass, which reticulate
## exposes as an R list.  The $ operator accesses its fields.
configs <- list(
  jm$make_config("model_lmm",          p = 2L),
  jm$make_config("model_lmmdriftdiff", p = 2L)
)
cat("model_lmm config:  free_drift =", configs[[1]]$free_drift,
    " free_diffusion =", configs[[1]]$free_diffusion, "\n")
cat("model_lmmdriftdiff: free_drift =", configs[[2]]$free_drift,
    " free_diffusion =", configs[[2]]$free_diffusion, "\n")
```

The two models differ in whether the SDE drift matrix **A** and diffusion matrix **G** are estimated:

| Model | Drift **A** | Diffusion **G** | Random slope **b** | Continuous-time SDE |
|---|:---:|:---:|:---:|---|
| `model_lmm` | fixed 0 | fixed 0 | free | Random slopes only; reduces to a linear mixed model |
| `model_lmmdriftdiff` | free | free | free | Full SDE with mean reversion and stochastic volatility |

# Cross-validated prediction

4-fold cross-validation is performed with a landmark time of 5 years; events are predicted over the window 5–15 years.  For each fold:

1. The LGSSM is fitted by MAP to the training individuals' longitudinal data.
2. The Kalman filter is run over **all** individuals.  For test-fold individuals, biomarker observations after the landmark time are withheld; the filter extrapolates the last known state using the SDE dynamics alone.
3. A Poisson GLM is fitted to the training data using the Kalman-filtered states as time-varying covariates.
4. Event probabilities are predicted for test-fold individuals over post-landmark intervals.

`jm$cross_validate()` orchestrates all four steps.  It accepts Python or R data frames transparently (reticulate converts R data frames to `pandas.DataFrame` before the call).  It returns a Python dict mapping model name to a `CVResult` object; reticulate converts this to an R named list.  Individual fields are accessed with `$`, just as with any R list.

```{r cv, message=FALSE}
## cross_validate() runs the full CV loop in Python.
## R data frames are accepted directly — reticulate converts them automatically.
## The result is a named R list; cv_results[["model_lmm"]]$pooled is an R list
## of summary statistics.
cv_results <- jm$cross_validate(
  dataSurv_split,
  dataLong_split,
  configs        = configs,
  biomarker_cols = biomarkers,
  timeinvar_surv = timeinvar_surv,
  landmark_time  = 5.0,
  n_folds        = 4L,
  random_seed    = 1234L,
  lgssm_steps    = 3000L,
  poisson_steps  = 3000L,
  verbose        = TRUE
)
```

# Results

## Kalman filter trajectories

The Kalman filter output is stored in the per-fold prediction data frames inside `cv_results`.  Each data frame contains the Kalman-filtered biomarker state means alongside the event and interval timing columns.  `py_to_r()` converts the Python data frame to an R data frame for plotting with `ggplot2`.

The plot shows the raw biomarker observations (points) and the Kalman-filtered state means (lines) for five individuals from the first cross-validation fold, using `model_lmmdriftdiff`.  Trajectories continue beyond the last observation date because the filter extrapolates using the fitted SDE.

```{r kalman_plot, fig.height=5}
## py_to_r() converts a pandas DataFrame to an R data frame.
## cv_results[["model_lmmdriftdiff"]]$fold_pred is an R list of data frames
## (one per fold); the first element is the fold-1 test predictions.
pred1    <- py_to_r(cv_results[["model_lmmdriftdiff"]]$fold_pred[[1]])
pred1_dt <- as.data.table(pred1)

## Select two individuals who had events and three who were censored.
ids_event <- head(pred1_dt[event == 1, unique(id)], 2)
ids_cens  <- head(setdiff(pred1_dt[event == 0, unique(id)], ids_event), 3)
ids_show  <- unique(c(ids_event, ids_cens))

kalman_long <- melt(
  pred1_dt[id %in% ids_show, .(id, tstart, logBili, albumin)],
  id.vars       = c("id", "tstart"),
  variable.name = "biomarker",
  value.name    = "filtered"
)

obs_long <- melt(
  as.data.table(dataLong_r)[id %in% ids_show, .(id, Time, logBili, albumin)],
  id.vars       = c("id", "Time"),
  variable.name = "biomarker",
  value.name    = "observed"
)

kalman_long[, id := factor(id)]
obs_long[,   id := factor(id)]

ggplot() +
  geom_line(data  = kalman_long,
            aes(x = tstart, y = filtered, colour = id),
            linewidth = 0.6) +
  geom_point(data = obs_long,
             aes(x = Time, y = observed, colour = id),
             size = 1.4, shape = 16) +
  facet_wrap(~ biomarker, scales = "free_y",
             labeller = labeller(biomarker = c(logBili = "log-bilirubin",
                                               albumin  = "albumin"))) +
  labs(x     = "Years from baseline",
       y     = "Standardised value",
       colour = "Individual",
       title  = "Kalman-filtered biomarker trajectories (model_lmmdriftdiff, fold 1)") +
  theme(legend.position = "right")
```

## Per-fold predictive performance

`cv_results[["model_lmm"]]$fold_stats` is a Python list of dicts, which reticulate converts to an R list of named lists.  Standard R functions (`mapply`, `sapply`, etc.) work directly on this structure.

```{r fold_table}
get_fold_table <- function(model_name) {
  stats_list <- cv_results[[model_name]]$fold_stats
  rows <- mapply(function(s, i) {
    data.frame(
      Fold          = i,
      Observed      = s[["Observed"]],
      Predicted     = round(s[["Predicted"]], 1),
      `Log score`   = round(s[["Log score"]], 2),
      `C-statistic` = round(s[["C-statistic"]], 3),
      check.names   = FALSE
    )
  }, stats_list, seq_along(stats_list), SIMPLIFY = FALSE)
  do.call(rbind, rows)
}

fold_lmm  <- get_fold_table("model_lmm")
fold_full <- get_fold_table("model_lmmdriftdiff")

kable(fold_lmm,
      caption = "model_lmm: per-fold predictive performance",
      align   = "r") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r fold_table2}
kable(fold_full,
      caption = "model_lmmdriftdiff: per-fold predictive performance",
      align   = "r") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Pooled cross-validated performance

```{r perf_table}
make_row <- function(model_name) {
  s <- cv_results[[model_name]]$pooled
  data.frame(
    Model          = model_name,
    Observed       = s[["Observed"]],
    Predicted      = round(s[["Predicted"]], 1),
    `Person-years` = round(s[["Person-years"]], 0),
    `Log score`    = round(s[["Log score"]], 2),
    `C-statistic`  = round(s[["C-statistic"]], 3),
    check.names    = FALSE
  )
}

perf_table <- rbind(make_row("model_lmm"),
                    make_row("model_lmmdriftdiff"))

kable(perf_table,
      caption = paste("Pooled 4-fold cross-validated predictive performance",
                      "(landmark time 5 years, follow-up to 15 years)"),
      align   = c("l", rep("r", 5))) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  column_spec(6, bold = TRUE)
```

## C-statistic by fold

```{r cstat_plot}
fold_df <- rbind(
  data.frame(model = "model_lmm",
             fold  = 1:4,
             C     = sapply(cv_results[["model_lmm"]]$fold_stats,
                            `[[`, "C-statistic")),
  data.frame(model = "model_lmmdriftdiff",
             fold  = 1:4,
             C     = sapply(cv_results[["model_lmmdriftdiff"]]$fold_stats,
                            `[[`, "C-statistic"))
)

ggplot(fold_df, aes(x = factor(fold), y = C, colour = model, group = model)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  geom_hline(data     = perf_table,
             aes(yintercept = `C-statistic`, colour = Model),
             linetype  = "dashed",
             linewidth = 0.5) +
  scale_y_continuous(limits = c(0.7, 1.0)) +
  labs(x      = "Fold",
       y      = "C-statistic",
       colour = "Model",
       title  = "C-statistic per fold (dashed line = pooled)") +
  theme(legend.position = "bottom")
```

The C-statistic and log-score are both computed over all person-time intervals from the landmark time to exit, treating each interval as an independent binary observation (event / no event).  The C-statistic is the area under the ROC curve of the predicted interval-event probability; the log-score is the sum of the Bernoulli log-likelihoods.

# Methods summary

The two-step procedure is:

1. **LGSSM** — a continuous-time SDE
   $$dX(t) = (AX(t) + b)\,dt + G\,dW(t), \quad y_t = x_t + v_t, \quad v_t \sim N(0,R)$$
   is fitted to the longitudinal data by MAP inference (NumPyro SVI with an `AutoDelta` guide and `ClippedAdam` optimiser).  Discrete-time transition matrices $A_d(\Delta t) = \exp(A\,\Delta t)$ and process noise $Q_d(\Delta t)$ are obtained via the Van Loan method, implemented with a fixed-order degree-7 Padé matrix exponential that contains no data-dependent conditionals and therefore compiles efficiently under JAX.

2. **Kalman filter** — implemented in JAX using `jax.lax.scan` over time steps and `jax.vmap` over individuals.  Filtered state means $\hat x_t$ are extracted for each person-time interval.  In the test fold, biomarker observations after the landmark time are withheld; the filter propagates the last observed state forward using only the SDE dynamics.

3. **Poisson GLM** — the log-hazard is modelled as
   $$\log \lambda_i(t) = \beta_0 + \beta_t t + \beta_x \hat x_i(t) + \beta_z z_i$$
   where $z_i$ are time-invariant covariates.  Fitted by MAP using QR decomposition of the design matrix for numerical stability.
