---
title: "jmseq: joint modelling of longitudinal and time-to-event data"
author: Paul McKeigue
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{jmseq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  cache     = TRUE,
  cache.path = "jmseq_cache/",
  fig.width  = 7,
  fig.height = 4
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

The `jmseq` package fits joint models of longitudinal biomarker trajectories and time-to-event outcomes.  The longitudinal submodel is a continuous-time linear Gaussian state space model (LGSSM) estimated by MAP inference via a JAX Kalman filter and NumPyro.  The survival submodel is a Poisson hazard model that uses the Kalman-filtered states as time-varying covariates.

This vignette demonstrates 4-fold cross-validated prediction of mortality on the Mayo Clinic primary biliary cirrhosis (PBC) dataset, comparing a linear mixed model (random slopes only) to a fuller SDE model with autoregressive drift and diffusion.

# Setup

```{r setup, cache=FALSE}
library(reticulate)
library(ggplot2)
library(kableExtra)
library(data.table)

use_python("/usr/bin/python3", required = TRUE)
ggplot2::theme_set(ggplot2::theme_light(base_size = 9))
```

```{r imports, cache=FALSE}
jm  <- import("jmseq")
np  <- import("numpy")
pd  <- import("pandas")
```

# Dataset

The PBC dataset comprises 1945 longitudinal measurements of two biomarkers (log-bilirubin and albumin, both standardised) on 312 individuals followed for mortality up to 15 years from baseline.  Time-invariant covariates are sex, age at baseline, and treatment assignment.

```{r data, cache=FALSE}
load(system.file("data", "pbc.rda", package = "jmseq"))

## convert trt factor to binary (D-penicillamine = 1)
dataSurv_r <- as.data.frame(pbc$dataSurv)
dataSurv_r$trt_binary <- as.integer(dataSurv_r$trt == "D-penicil")
dataSurv_r <- dataSurv_r[, c("id", "sex", "agebaseline", "trt_binary",
                              "Time.cens", "event")]

dataLong_r <- as.data.frame(pbc$dataLong[, c("id", "Time",
                                              "logBili", "albumin")])

biomarkers     <- c("logBili", "albumin")
timeinvar_surv <- c("sex", "agebaseline", "trt_binary")

cat(sprintf("dataSurv: %d individuals, %d events\n",
            nrow(dataSurv_r), sum(dataSurv_r$event)))
cat(sprintf("dataLong: %d observations\n", nrow(dataLong_r)))
```

## Preprocessing: interval splitting

Follow-up time is split into intervals of at most 0.25 years.  Biomarker observations recorded within each interval are used to update the Kalman filter state.

```{r split}
surv_py <- r_to_py(dataSurv_r)
long_py <- r_to_py(dataLong_r)

split_df <- jm$split_surv_long(
  surv_py,
  long_py,
  max_interval  = 0.25,
  timeinvar_surv = timeinvar_surv,
  biomarkers     = biomarkers
)

dataSurv_split <- split_df[c("id", "Time", "tstop", "event",
                              timeinvar_surv)]
dataLong_split <- split_df[c("id", "Time", "tstop", biomarkers)]

cat(sprintf("After splitting: %d rows, max T per individual = %d\n",
            nrow(split_df),
            max(table(py_to_r(split_df["id"])))))
```

# Model variants

We compare two LGSSM variants for the biomarker trajectories.

```{r configs}
configs <- list(
  jm$make_config("model_lmm",          p = 2L),
  jm$make_config("model_lmmdriftdiff", p = 2L)
)
```

The two models differ in whether the SDE drift matrix **A** and diffusion matrix **G** are estimated:

| Model | Drift **A** | Diffusion **G** | Random slope **b** |
|---|---|---|---|
| `model_lmm` | fixed 0 | fixed 0 | free |
| `model_lmmdriftdiff` | free | free | free |

# Cross-validated prediction

4-fold cross-validation with landmark time 5 years.  For each fold the LGSSM is fitted by MAP to the training data, the Kalman filter is run over all individuals, and a Poisson GLM is fitted using the filtered states as time-varying predictors of the hazard.  Predictive performance is evaluated on the test fold over the window 5–15 years.

```{r cv, message=FALSE}
cv_results <- jm$cross_validate(
  dataSurv_split,
  dataLong_split,
  configs        = configs,
  biomarker_cols = biomarkers,
  timeinvar_surv = timeinvar_surv,
  landmark_time  = 5.0,
  n_folds        = 4L,
  random_seed    = 1234L,
  lgssm_steps    = 3000L,
  poisson_steps  = 3000L,
  verbose        = FALSE
)
```

## Kalman filter trajectories

The plot below shows the raw biomarker observations (points) and the Kalman-filtered state means (lines) for five individuals from the first cross-validation fold, using `model_lmmdriftdiff`.

```{r kalman_plot, fig.height=5}
## Extract fold-1 prediction data (contains Kalman-filtered biomarker columns)
pred1 <- py_to_r(cv_results[["model_lmmdriftdiff"]]$fold_pred[[1]])
pred1_dt <- as.data.table(pred1)

## Take five representative individuals (first two who died, first three who were censored)
ids_event <- head(pred1_dt[event == 1, unique(id)], 2)
ids_cens  <- head(setdiff(pred1_dt[event == 0, unique(id)], ids_event), 3)
ids_show  <- unique(c(ids_event, ids_cens))

kalman_long <- melt(
  pred1_dt[id %in% ids_show, .(id, tstart, logBili, albumin)],
  id.vars    = c("id", "tstart"),
  variable.name = "biomarker",
  value.name    = "filtered"
)

obs_long <- melt(
  as.data.table(dataLong_r)[id %in% ids_show, .(id, Time, logBili, albumin)],
  id.vars    = c("id", "Time"),
  variable.name = "biomarker",
  value.name    = "observed"
)

kalman_long[, id := factor(id)]
obs_long[, id := factor(id)]

ggplot() +
  geom_line(data  = kalman_long,
            aes(x = tstart, y = filtered, colour = id),
            linewidth = 0.6) +
  geom_point(data = obs_long,
             aes(x = Time, y = observed, colour = id),
             size = 1.2, shape = 16) +
  facet_wrap(~ biomarker, scales = "free_y",
             labeller = labeller(biomarker = c(logBili = "log-bilirubin",
                                               albumin  = "albumin"))) +
  labs(x = "Years from baseline", y = "Standardised value",
       colour = "Individual",
       title  = "Kalman-filtered biomarker trajectories (model_lmmdriftdiff, fold 1)") +
  theme(legend.position = "right")
```

## Per-fold predictive performance

```{r fold_table}
get_fold_table <- function(model_name) {
  stats_list <- cv_results[[model_name]]$fold_stats
  rows <- mapply(function(s, i) {
    data.frame(
      Fold           = i,
      Observed       = s[["Observed"]],
      Predicted      = round(s[["Predicted"]], 1),
      `Log score`    = round(s[["Log score"]], 2),
      `C-statistic`  = round(s[["C-statistic"]], 3),
      check.names    = FALSE
    )
  }, stats_list, seq_along(stats_list), SIMPLIFY = FALSE)
  do.call(rbind, rows)
}

fold_lmm  <- get_fold_table("model_lmm")
fold_full <- get_fold_table("model_lmmdriftdiff")

kable(fold_lmm,
      caption = "model_lmm: per-fold predictive performance",
      align   = "r") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r fold_table2}
kable(fold_full,
      caption = "model_lmmdriftdiff: per-fold predictive performance",
      align   = "r") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Pooled cross-validated performance

```{r perf_table}
make_row <- function(model_name) {
  s <- cv_results[[model_name]]$pooled
  data.frame(
    Model         = model_name,
    Observed      = s[["Observed"]],
    Predicted     = round(s[["Predicted"]], 1),
    `Person-years` = round(s[["Person-years"]], 0),
    `Log score`   = round(s[["Log score"]], 2),
    `C-statistic` = round(s[["C-statistic"]], 3),
    check.names   = FALSE
  )
}

perf_table <- rbind(make_row("model_lmm"),
                    make_row("model_lmmdriftdiff"))

kable(perf_table,
      caption = paste("Pooled 4-fold cross-validated predictive performance",
                      "(landmark time 5 years, follow-up to 15 years)"),
      align   = c("l", rep("r", 5))) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  column_spec(6, bold = TRUE)
```

## C-statistic by fold

```{r cstat_plot}
fold_df <- rbind(
  data.frame(model = "model_lmm",          fold = 1:4,
             C = sapply(cv_results[["model_lmm"]]$fold_stats,
                        `[[`, "C-statistic")),
  data.frame(model = "model_lmmdriftdiff", fold = 1:4,
             C = sapply(cv_results[["model_lmmdriftdiff"]]$fold_stats,
                        `[[`, "C-statistic"))
)

ggplot(fold_df, aes(x = factor(fold), y = C, colour = model, group = model)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  geom_hline(data = perf_table,
             aes(yintercept = `C-statistic`, colour = Model),
             linetype = "dashed", linewidth = 0.5) +
  scale_y_continuous(limits = c(0.7, 1.0)) +
  labs(x = "Fold", y = "C-statistic",
       colour = "Model",
       title  = "C-statistic per fold (dashed = pooled)") +
  theme(legend.position = "bottom")
```

# Methods summary

The two-step procedure is:

1. **LGSSM** — a continuous-time SDE
   $$dX(t) = (AX(t) + b)\,dt + G\,dW(t), \quad y_t = x_t + v_t, \quad v_t \sim N(0,R)$$
   is fitted to the longitudinal data by MAP inference (NumPyro SVI with an `AutoDelta` guide).  Discrete-time transition matrices $A_d(\Delta t) = \exp(A\,\Delta t)$ and process noise $Q_d(\Delta t)$ are obtained via the Van Loan method.

2. **Kalman filter** — the filtered state means $\hat x_t$ are extracted for each person-time interval.  In the test fold, biomarker observations after the landmark time are withheld; the filter propagates the last observed state forward using only the SDE dynamics.

3. **Poisson GLM** — the log-hazard is modelled as
   $$\log \lambda_i(t) = \beta_0 + \beta_t t + \beta_x \hat x_i(t) + \beta_z z_i$$
   where $z_i$ are time-invariant covariates.  Fitted with QR decomposition for numerical stability.

The **C-statistic** and **log-score** are both computed over all person-time intervals from the landmark time to exit, treating each interval as an independent binary observation.
